Ultrafast Control of Spin-Spin Interactions in Two-Dimensional Magnets

### 2019-03-10 Paper/paper.tex
v6

\begin{abstract}
Light provides ultrafast control of exchange and Dzyaloshinskii-Moriya interactions, we consider two-dimensional honeycomb lattices described by the Kane-Mele-Hubbard model at half-filling and in the strongly correlated limit, i.e., the Mott insulator regime of a canted antiferromagnet. Based on the Floquet theory, we demonstrate that changing the amplitude and frequency of circularly polarized laser pulses, tunes the amplitude, the sign and even the ratio between these two spin interactions. Furthermore, the renormalizations of the spin interactions are helicity independent. Our results pave the way for ultrafast optical manipulation of spins in recently discovered two-dimensional magnetic materials.
\end{abstract}

@@

\textit{Introduction}.--- The discovery of all-optical control of the order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by ultrashort, intense and high-frequency laser pulses has transformed spintronics to a new era of ultrafast magnetism \cite{kimel2005ultrafast,Rasing}. Although the underlying microscopic mechanism remains unclear, there have been many attempts to uncover its origins. Recently, it has been theoretically proposed that laser pulses can modify and even reverse the sign of the exchange interaction \cite{PhysRevLett.113.057201,PhysRevLett.115.075301,Mentink2015,PhysRevLett.118.097701,Stepanov2017,Kitamura,MentinkReview,Barbeau2018}, the strongest spin interaction in magnetic systems. This finding has been supported by recent experiments \cite{2018Natur.553..481G,Mikhaylovskiy2015,Perakis,PhysRevLett.93.197403}. Another important spin interaction is the Dzyaloshinskii-Moriya interaction (DMI), an antisymmetric exchange interaction in inversion symmetry broken systems, where the chiral symmetry of the system is broken \cite{dzyaloshinsky1958, moriya1960,Alireza-DMI1, Alireza-DMI2}. While this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials and causes weak ferromagnetism in AFM materials \cite{dzyaloshinsky1958, moriya1960}, topological objects like chiral skyrmions \cite{2009Sci...323..915M,2017NatRM...217031F,AlirezaSk1, AlirezaSk2} and chiral domain walls \cite{Thiaville_2012,Parkin,AlirezaSW}, and exotic phase of topological magnon insulators \cite{Owerre2016, Elyasi2018, Chen2018, Owerre1, Owerre2}. The ratio between these two interactions controls the tilting angle of the canted spins. Finding a mechanism to tune this ratio enables new phenomena in ultrafast spin dynamics and switching \cite{Mikhaylovskiy2015,Mentink2015,Stepanov2017,Kitamura}.

@@

Using the variational principle, it has been shown that the NN Coulomb interaction can be approximated by a renormalized local interaction $U =U_{00} - \bar{V}$, where $\bar{V}$ is a weighted average of the NN Coulomb interaction \cite{Schuler2013,Stepanov2017}. Thus, we only take into account the local Coulomb interaction in our total Hamiltonian and express the interaction part of the Hamiltonian in Eq. (\ref{Hint}) as
\begin{equation}
\hat{\mathcal{H}}_{\text{int}} \approx U\hat{d},
\end{equation}
where we define the doublon number operator $\hat{d} = \sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$ with eigenvalues $d$. We denote the projection operator to the subspace of states with eigenvalue $d$ (i.e. states with exactly $d$ double occupancies) by $\hat{P}_d$. We are interested in the strongly correlated regime $U\gg t_{1(2)}$. In such a strong-coupling limit any state with a nonzero number of double occupancies ($d \neq 0$) has a much larger energy than those with no double occupancies ($d=0$). We obtain an effective Hamiltonian acting on the $d=0$ subspace by standard second-order perturbation theory in the hopping terms. Using the following relations,
\begin{align}
&\hat{c}_{i \tau}^\dagger \hat{c}_{i \tau'} = \frac{1}{2} (n_{i \uparrow} + n_{i \downarrow})\delta_{\tau \tau'}  + \bs{S}_i\cdot\bs{\tau}_{\tau', \tau}, \label{SpinOperatorInv1}\\
&\hat{c}_{i \tau} \hat{c}_{i \tau'}^\dagger = \frac{1}{2} (2 - n_{i \uparrow} - n_{i \downarrow}) \delta_{\tau \tau'} - \bs{S}_i\cdot\bs{\tau}_{\tau, \tau'}, \label{SpinOperatorInv2}
\end{align}
we find the spin Hamiltonian for 2D AFM Mott insulators,
\begin{align}
\label{MKMHeff0}
H_{\text{S}} =& J_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + J_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j,  \n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \bs{\Gamma}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \bs{D}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following spin-spin interactions,
\begin{subequations}
\label{spin-para}
\begin{align}
&J_{1(2)} = \frac{2t_{1(2)}^2}{U}, \\
&\bs{\Gamma}_{ij} =\frac{2\Delta^2}{U} \text{diag}(-1,-1,1),\\
&\bs{D}_{ij} = \frac{4 t_2 \Delta}{U}\nu_{ij}  \hat{\mathrm{e}}_z.
\end{align}
\end{subequations}
The first and second terms in the spin Hamiltonian, Eq. (\ref{MKMHeff0}), are the NN and NNN symmetric Heisenberg AFM exchange interactions, $J_{1(2)}>0$, respectively; the third term is the NNN anisotropic Heisenberg exchange interaction (XXZ-like term) arising from the intrinsic SOI, and finally, the last term is the intrinsic NNN DMI. The intrinsic SOI in Eq. (\ref{Hsoi}), leads to a DMI vector, $\bs{D}_{ij}$, perpendicular to the honeycomb plane. It can also be shown that the Rashba SOI results in an in-plane DMI vector that is perpendicular to the lattice bonds.
For the sake of completeness, we explain shortly the effect of disorder by adding an on-site disorder potential $\sum_{i \tau} \varepsilon_i \hat{c}_{i \tau}^\dagger \hat{c}_{i \tau}$ in the Kane-Mele-Hubbard Hamiltonian Eq. (\ref{MKMH}), where $\varepsilon_i$ is an uncorrelated random variable. Following the above procedure, it can be shown that the spin interaction parameters are renormalized as $U^{-1} \rightarrow U/\left(U^2-(\epsilon_j-\epsilon_i)^2\right)$ \cite{Protopopov2018}. In the large $U$ limit, the effect of disorder is negligible and we thus ignore it in the rest of this Letter.

>>>>

\begin{abstract}
Light provides ultrafast, direct and non-thermal control of exchange and Dzyaloshinskii-Moriya interactions. We consider two-dimensional honeycomb lattices described by the Kane-Mele-Hubbard model at half-filling and in the strongly correlated limit, i.e., the Mott insulator regime of a canted antiferromagnet. Based on the Floquet theory, we demonstrate that changing the amplitude and frequency of polarized laser pulses, tunes the amplitude, the sign and even the ratio between these two spin interactions. Furthermore, the renormalizations of the spin interactions are helicity independent. Our results pave the way for ultrafast optical manipulation of spins in recently discovered two-dimensional magnetic materials.
\end{abstract}

@@

\textit{Introduction}.--- The discovery of all-optical control of the order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by ultrashort, intense and high-frequency laser pulses has transformed spintronics to a new era of ultrafast magnetism \cite{kimel2005ultrafast,Rasing}. 
Recent experiments have shown that laser pulses can directly modify the amplitude and sign of the exchange interaction, the strongest spin interaction in magnetically ordered systems \cite{2018Natur.553..481G,Mikhaylovskiy2015,Perakis,PhysRevLett.93.197403}. Although the underlying microscopic mechanism remains unclear, there have been many attempts to uncover its origins.   
Recently, it has been theoretically proposed that the non-thermal optical modification of exchange interaction is possible through direct coupling between electric field of light and spins \cite{PhysRevLett.113.057201,PhysRevLett.115.075301,Mentink2015,PhysRevLett.118.097701,Stepanov2017,Kitamura,MentinkReview,Barbeau2018}.
In magnetic systems with broken inversion symmetry there is an antisymmetric exchange interaction between spins that breaks the chiral symmetry called the Dzyaloshinskii-Moriya interaction (DMI) \cite{dzyaloshinsky1958, moriya1960,Alireza-DMI1, Alireza-DMI2}. Although this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials and causes weak ferromagnetism in AFM materials \cite{dzyaloshinsky1958, moriya1960}, topological objects like chiral skyrmions \cite{2009Sci...323..915M,2017NatRM...217031F,AlirezaSk1, AlirezaSk2} and chiral domain walls \cite{Thiaville_2012,Parkin,AlirezaSW}, and exotic phase of topological magnon insulators \cite{Owerre2016, Elyasi2018, Chen2018, Owerre1, Owerre2}. The ratio between these two interactions controls the tilting angle of the canted spins. Finding a mechanism to tune this ratio enables new phenomena in ultrafast spin dynamics and switching \cite{Mikhaylovskiy2015,Mentink2015,Stepanov2017,Kitamura}.

@@

Using the variational principle, it has been shown that the NN Coulomb interaction can be approximated by a renormalized local interaction $U =U_{00} - \bar{V}$, where $\bar{V}$ is a weighted average of the NN Coulomb interaction \cite{Schuler2013,Stepanov2017}. Thus, we only take into account the local Coulomb interaction in our total Hamiltonian and express the interaction part of the Hamiltonian in Eq. (\ref{Hint}) as
\begin{equation}
\hat{\mathcal{H}}_{\text{int}} \approx U\hat{d},
\end{equation}
where we define the doublon number operator $\hat{d} = \sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$ with eigenvalues $d$. We denote the projection operator to the subspace of states with eigenvalue $d$, i.e., states with exactly $d$ double occupancies, by $\hat{P}_d$. We are interested in the strongly correlated regime $U\gg t_{1(2)}$ at half-filling. In such a strong-coupling limit any state with a nonzero number of double occupancies ($d \neq 0$) has a much larger energy than those with no double occupancies ($d=0$). We obtain an effective Hamiltonian acting on the $d=0$ subspace by standard second-order perturbation theory in the hopping terms. Using the following relations,
\begin{align}
&\hat{c}_{i \tau}^\dagger \hat{c}_{i \tau'} = \frac{1}{2} (n_{i \uparrow} + n_{i \downarrow})\delta_{\tau \tau'}  + \bs{S}_i\cdot\bs{\tau}_{\tau', \tau}, \label{SpinOperatorInv1}\\
&\hat{c}_{i \tau} \hat{c}_{i \tau'}^\dagger = \frac{1}{2} (2 - n_{i \uparrow} - n_{i \downarrow}) \delta_{\tau \tau'} - \bs{S}_i\cdot\bs{\tau}_{\tau, \tau'}, \label{SpinOperatorInv2}
\end{align}
we find the spin Hamiltonian for 2D AFM Mott insulators,
\begin{align}
\label{MKMHeff0}
H_{\text{S}} =& J_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + J_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j,  \n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \bs{\Gamma}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \bs{D}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following spin-spin interactions,
\begin{subequations}
\label{spin-para}
\begin{align}
&J_{1(2)} = \frac{2t_{1(2)}^2}{U}, \\
&\bs{\Gamma}_{ij} =\frac{2\Delta^2}{U} \text{diag}(-1,-1,1),\\
&\bs{D}_{ij} = \frac{4 t_2 \Delta}{U}\nu_{ij}  \hat{\mathrm{e}}_z.
\end{align}
\end{subequations}
The first and second terms in the spin Hamiltonian, Eq. (\ref{MKMHeff0}), are the NN and NNN symmetric Heisenberg AFM exchange interactions, $J_{1(2)}>0$, respectively; the third term is the NNN anisotropic Heisenberg exchange interaction (XXZ-like term) arising from the intrinsic SOI, and finally, the last term is the intrinsic NNN DMI. The intrinsic SOI in Eq. (\ref{Hsoi}), leads to a DMI vector, $\bs{D}_{ij}$, perpendicular to the honeycomb plane. It can also be shown that the Rashba SOI results in an in-plane DMI vector that is perpendicular to the lattice bonds.
For the sake of completeness, we explain shortly the effect of disorder by adding an on-site disorder potential $\sum_{i \tau} \varepsilon_i \hat{c}_{i \tau}^\dagger \hat{c}_{i \tau}$ in the Kane-Mele-Hubbard Hamiltonian Eq. (\ref{MKMH}), where $\varepsilon_i$ is an uncorrelated random variable. Following the above procedure, it can be shown that the spin interaction parameters are renormalized as $U^{-1} \rightarrow U/\left(U^2-(\epsilon_j-\epsilon_i)^2\right)$ \cite{Protopopov2018}. In the large $U$ limit, the effect of disorder is negligible and we thus ignore it in the rest of this Letter.
